State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-04-15 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
paper edits

Due to the high dimensionality of the brain and the current electrode systems, we can not expect to have more sensors than basis functions.
\section{State and Parameter Estimation}
The estimation process is a two part iterative algorithm that yields the estimate of the connectivity kernel parameters and the posterior distribution of the field weights. In the parameter estimation step, Least Squares (LS) method is used which minimizes the sum of the squares of the errors on each state equation (see Appendix). In the state estimation step, given an estimate of $\boldsymbol \theta$ and $\xi$ the Unscented Rauch Tung Striebel Smoother (URTSS) \cite{Sarkka2010} was used to determine the posterior state sequence .
The (URTSS) uses the Unscented Kalman Filter (UKF) in the forward iteration \cite{Julier1997} to calculate a forward posterior estimate $\hat{\mathbf x}_t^{f} $ along with a separate backward smoothing path to compute the smooth estimations $\hat{\mathbf x}_t^{b} $. The method is appropriate for state estimation of stochastic nonlinear dynamical systems. Alternatives to the UKF are the Extended Kalman Filter (EKF) \cite{Haykin2001}  and Sequential Monte Carlo (SMC) filtering \cite{doucet2001}. The extended Kalman filter approximates the state transition equation by linearizing about the fixed points (by calculating the Jacobian) at each time step. The linearization maintains the Gaussianality in the model. The unscented Kalman filter approximates the \emph{a posteriori} state density by a Gaussian distribution using a minimal set of carefully chosen sigma points, whilst maintaining the nonlinearity in the system. This has been shown to give superior performance over the EKF in state estimation, as the EKF maintains a first order approximation where the UKF provides an approximation accurate at least to the second order. In addition, the computation of the Jacobian used in the
EKF can be problematic. SMC filtering can theoretically provide an exact \emph{a posteriori} state density for nonlinear systems. However, currently this method is not appropriate for our problem, due to computation demands.

@@

Propogate the sigma points through the state equation

>>>>

Due to the high dimensionality of the brain and the current electrode systems, we can not expect to have more sensors than basis functions.
\section{State and Parameter Estimation}\label{StateAndParameterEstimationSection}
In this section we describe the procedure for estimating the states that scale the field basis functions, $\mathbf{x}_t$, the connectivity kernel parameters, $\boldsymbol \theta$, and the synaptic time constant, $\xi$. The estimation process consists of a two part iterative algorithm, consisting of the state estimation step followed by the parameter estimation step at each time instance. Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation utilises a least squares (LS) method that minimises the sum of the squared errors (of a predicted state update) with each new state estimate (see Appendix~\ref{LeastSquaresAppendix}).

The state estimation step utilises the Unscented Rauch Tung Striebel Smoother (URTSS)~\cite{Sarkka2010} to determine the posterior state sequence. The (URTSS) uses the Unscented Kalman Filter (UKF) in the forward iteration~\cite{Julier1997} to calculate a forward posterior state estimate, $\hat{\mathbf x}_t^{f}$, followed by a backward path to compute the smoothed state estimations $\hat{\mathbf x}_t^{b}$. The method is appropriate for state estimation of stochastic nonlinear dynamical systems. Alternatives to the UKF are the Extended Kalman Filter (EKF) \cite{Haykin2001}  and Sequential Monte Carlo (SMC) filtering \cite{doucet2001}. The extended Kalman filter approximates the state transition equation by linearizing about the fixed points (by calculating the Jacobian) at each time step. The linearization maintains the Gaussianality in the model. The unscented Kalman filter approximates the \emph{a posteriori} state density by a Gaussian distribution using a minimal set of carefully chosen sigma points, whilst maintaining the nonlinearity in the system. This has been shown to give superior performance over the EKF in state estimation, as the EKF maintains a first order approximation where the UKF provides an approximation accurate at least to the second order. In addition, the computation of the Jacobian used in the
EKF can be problematic. SMC filtering can theoretically provide an exact \emph{a posteriori} state density for nonlinear systems. However, currently this method is not appropriate for our problem, due to computation demands.

@@

Propagate the sigma points through the state equation
